A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
switch added

@@

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

Our final algorithm is quite technical and has numerous moving parts. This section
provides a high level view of the entire result, and highlights the main ideas
and intuition. 
It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
terrain. In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body. They may also differ from the original definitions proposed
in earlier work.

@@

\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
to design an algorithm for contour trees on such inputs. 
For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
except at a unique boundary face (think of a large boundary triangle that is the boundary).
All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.

@@

%\section{Acknowledgements}
\paragraph{Acknowledgements.}
We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy’s National
Nuclear Security Administration under contract DE-AC04-94AL85000.

\section{Divide and conquer through contour surgery} \label{sec:surgery}

@@

% \section{Temp}
% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
% 
% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
% 
% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
% Hence we can assume no such edges exist.  Now we have three cases:
% 
% \begin{enumerate}[{Case} 1]
%  \item 
% $l$ is adjacent to a min path:\newline
% Note in this case the path containing $u$ must be a min path.
% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
% \item
% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
% \item
% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a join).
% \end{enumerate}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def \STOC {}

\ifx\STOC\undefined
  \newcommand{\FullVer}[1]{#1}
  \newcommand{\SubmitVer}[1]{}
\else
  \newcommand{\FullVer}[1]{}
  \newcommand{\SubmitVer}[1]{#1}
\fi

@@

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
specific path decomposition $P(T)$%
\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

Our final algorithm is quite technical and has numerous moving parts. This section
provides a high level view of the entire result, and highlights the main ideas
and intuition. 
\SubmitVer{The full version with all the details can be found in the appendix.}
It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
terrain. In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body. They may also differ from the original definitions proposed
in earlier work.

@@

\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
to design an algorithm for contour trees on such inputs. 
For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
except at a unique boundary face (think of a large boundary triangle that is the boundary).
All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.

@@

\FullVer{
\section{Divide and conquer through contour surgery} \label{sec:surgery}

@@

% \section{Temp}
% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
% 
% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
% 
% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
% Hence we can assume no such edges exist.  Now we have three cases:
% 
% \begin{enumerate}[{Case} 1]
%  \item 
% $l$ is adjacent to a min path:\newline
% Note in this case the path containing $u$ must be a min path.
% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
% \item
% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
% \item
% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a join).
% \end{enumerate}
}

\paragraph{Acknowledgements.}
We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy’s National
Nuclear Security Administration under contract DE-AC04-94AL85000.
